Linear response theory and optimal control for a molecular system under nonequilibrium conditions

### 2013-08-21 writedown/linear.response/paper.tex
Carsten's revise, 2nd round

\begin{abstract}
  In this paper, we propose a straightforward generalization of linear response theory on a finite time-horizon to systems in nonequilibrium that are subject to external forcing driving.  We briefly revisit the standard linear response result for equilibrium systems, where we consider Langevin dynamics as a special case, and then give an alternative derivation using a change-of-measure argument that does not rely on any stationarity or reversibility assumption. This procedure easily enables us to calculate the second order correction to the linear response formula (which may or may not be useful in practice). Furthermore, we outline how the novel nonequilibrium linear response formula can be used to compute optimal controls of molecular systems for cases in which one wants to steer the system to maximize a certain target expectation value. We illustrate our approach with simple numerical examples. 
\end{abstract}

@@

Often, however, one is interested in knowing about the response of the molecular system to perturbation out of equilibrium. The standard linear response formula allows to answer this question, at least partially. In the standard setting it gives us the first order of the change to an equilibrium expectation value as resulting from the nonequilibrium perturbation. Here, first order means first order in the size of the perturbation. This linear response formula has a long history of extensions and generalizations; see, e.g., \cite{marconi2008,ruelle2009} and the references therein. In some sense it has become one of the cornerstones of modern statistical physics since it can be related to the fluctuation dissipation theorem (FDT) which roughly states that for appropriate systems in statistical equilibrium, the average response to small external perturbations can be calculated through the knowledge of suitable correlation functions of the unperturbed statistical system. Linear response formulae on the infinite time-horizon have been shown to hold under rather mild assumptions and for nonequilibrium systems with a unique invariant measure \cite{hairer2008}.

There is an increasing number of articles in the literature that report on applications of molecular dynamics to nonequilibrium settings. There are many generalization to so-called nonequilibrium steady states based on the generality of the FDT \cite{Seifert}, but despite its wide use, we are not aware of a linear response formula, even for nonequilibrium (i.e.~irreversible) systems, that does not rely on any kind of stationarity assumption. We will provide such a formula for the case that the underlying dynamics can be described by Langevin dynamics that may not have an invariant measure. Furthermore, we will provide a formula for the second order response of a nonequilibrium Langevin system subject to a small perturbation.

Instead of applying this theory to a molecular system we will go one step further. We will outline how the novel nonequilibrium linear response formula can be used to compute the optimal control of molecular systems. In optimal control one seeks the optimal way to perturb a molecular system such that a certain target expectation value (e.g. population of certain states) is maximized under constraints on the energy of the control. In general the control drives the molecular system under control out of equilibrium. Thus, the nonequilibrium linear response formula can be used to find the optimal correction of the present control regarding the expectation value of interest.

@@

In this section, we first want to give a simple and formal derivation of the linear response for equilibrium and nonequilibrium systems. To this end we consider a general It\^o stochastic differential equation of the form
\begin{equation}\label{sde}
dX_{t} = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t}\,,\quad t\ge 0\,,
\end{equation}
where $X_{t}\in\R^{d}$, $b(\cdot,\cdot)$ is a smooth time-dependent vector field, $a(\cdot)$ a smooth field of $d\times n$ matrices and $B_{t}$ is standard Brownian motion in $\R^{n}$. Here $v_{t}\in\R^{d}$ is any given driving force applied to the system, typically an affine function of the form $v_{t}=B(X_{t})u(t)$, with $B(\cdot)\in\R^{d\times k}$ and a bounded measurable function $u\colon[0,\infty)\to\R^{k}$ such that (\ref{sde}) has a strong solution for all $t> 0$ whenever $\eps>0$ is sufficiently small. When (\ref{sde}) without the perturbation $\eps v$ is considered, we write
\begin{equation}\label{sdewo}
dx_{t} = b(x_{t},t)dt +a(x_{t})dB_{t}\,,\quad t\ge 0\,. 
\end{equation}

\subsection{Small perturbations from equilibrium: Langevin dynamics}

@@

Specifically, we are interested in the case that (\ref{sde}) has the form of a second-order Langevin equation, in which case $d=2n$ and 
\begin{equation}\label{eqn:langevin-1}
b(x,t) = (J-R)\nabla H(x)\,,\quad a = \sqrt{2\beta^{-1}}R^{1/2}\,,
\end{equation}
with $J=-J^{T}$ the canonical $2n\times 2n$ symplectic matrix,
\[
J = \left ( \begin{array}{cc} 
0 & I_{n\times n} \\ -I_{n\times n} &0
\end{array}\right),
\]
$R=R^{T}\ge 0$ the positive semidefinite $2n\times 2n$ friction matrix 
\[
R = \left ( \begin{array}{cc}
0 & 0 \\ 0 & \gamma
\end{array}\right),
\]
for $\gamma\in\R^{n\times n}$ symmetric positive definite, and 
\[
H\colon\R^{2n}\to\R\,,\quad H(x) = \frac{1}{2}|x_{2}|^{2} + V(x_{1})
\] 
the Hamiltonian of the system. We assume that the potential $V$ is bounded from below. Then, under some mild growth conditions on $V$ for $|x_{1}|\to\infty$ (see, e.g., \cite{mattingly2002}), the unperturbed dynamics has a unique invariant measure with density 
\[
\rho_{\infty}(x) = \frac{1}{Z}\exp(-\beta H(x))\,,\quad  Z=\int_{\R^{2n}}  \exp(-\beta H(x))\,dx\,.
\] 
In this case, the linear response result (\ref{lr}) can be recast in form of the better known Green-Kubo formula \cite{risken1996}. Assuming that the forcing is of the form $v_{t}=Bu(t)$ with $u\colon[0,\infty)\to\R^{n}$ and $B\in\R^{2n\times n}$ being any given time-dependent function, the above expression can be recast as 
\begin{equation}\label{GreenKubo}
\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f] - \bE_{\rho^{0}}[f]}{\eps} = \beta\int_{0}^{t}F(t-s)u(s)\,ds
\end{equation}
where 
\[
F(t) = - \int_{\R^{2n}}\bE_{x}[f(x_{t})] B^{T}\nabla H(x)\rho_{\infty}(x)\,dx
\]
is the response function and the expectation under the integral is taken over all realization of the equilibrium process (\ref{sdewo}) starting from $x_{0}=x$. In other words: 
\[
\bE_{\rho^{\eps}}[f] \approx \bE_{\rho_{\infty}}[f] + \eps\beta\int_{0}^{t}F(t-s)u(s)\,ds\,,
\]
where we have used the fact that $\rho^{0}(x,t)\to \rho_{\infty}(x)$ uniformly on $\R^{2n}\times[0,\infty)$.

@@

\subsubsection*{Girsanov transformation}
We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal2003stochastic}. To this end, let $x_{t}=x_{t}(\omega)$ and $X_{t}=X_{t}(\omega)$ be the solutions of our generic stochastic differential equations
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t}\,,\quad 0\le t\le T  \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t}\,,\quad 0\le t\le T \label{sde2-2}
\end{align}
\end{subequations}
for $T<\infty$ and deterministic initial conditions
\[
x_{0}(\omega) = X_{0}(\omega) = x\quad\textrm{(almost surely)}\,.
\]
Suppose that there exists an auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,.
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to rewrite (\ref{sde2-2}) as
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003stochastic}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{stroock2006}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$. Specifically, let $\nu$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $\mu$ on the space of continuous trajectories by 
\[
d\mu=M_{T}d\nu
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any functional 
\[
\varphi(\omega)=\int_{0}^{T}f((X_{s}(\omega))\,ds + g(X_{T}(\omega))
\] 
that is integrable \wrt $\nu$, we have the identity\footnote{One of the omitted technical details is Novikov's condition \cite[pp.~162]{oksendal2003stochastic} that guarantees that $(M_{t})_{0\le t\le T}$ is a Martingale, which by $\bE_{\mu}[1]=\bE_{\nu}[M_{T}]=1$ implies normalizability of the new probability measure $\mu$.}
\[
\bE_{\nu}[\varphi] := \int_{\Omega} \varphi(\omega)\, d\nu(\omega) = \int_{\Omega} \varphi(\omega) M_{T}^{-1}(\omega)\,d\mu(\omega) =: \bE_{\mu}[M^{-1}_{T}\varphi]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)
\]
is the density of $\nu$ relative to $\mu$. Note that the expectation on the right hand side corresponds to the unperturbed dynamics, because $W_{t}$ is a standard Brownian motion under $\mu$, and the expectation is over all realizations of (\ref{sde2-3}) starting from either any given initial condition $X_{0}=x$ or an arbitrary initial distribution. On the other hand, $X_{t}$ under $\nu$ corresponds to the perturbed dynamics, which should become clear upon comparing equations (\ref{sde2-2}) and (\ref{sde2-3}).

@@

Linearization of $M_{T}^{-1}$ about $\eps=0$, assuming that the control has bounded variance, yields the alternative linear response formula 
\begin{equation}\label{lr-alt}
\lim_{\eps\to 0}\frac{\bE_{\rho_{x}^{\eps}}[\varphi] + \bE_{\rho_{x}^{0}}[\varphi]}{\eps} = \bE_{\rho_{x}^{0}}\left[\varphi \int_{0}^{T}\xi_{s}\cdot dB_{s} \right]\,,
\end{equation}
where, in accordance with the previous linear response formula, $\rho_{x}^{\eps}$ and $\rho_{x}^{0}$ denote the path probability distributions of perturbed and unperturbed processes (\ref{sde2-2}) and (\ref{sde2-1}) with deterministic initial condition. On the right hand side of the last equation $\varphi$ is understood as a functional of $(x_{t})_{0\le t\le T}$, the solution to (\ref{sde2-1}) with initial condition $x_{0}=x$, and $\xi_{t}$ is the solution to $a(x_{t})\xi_{t}=v_{t}$. By averaging the the initial values $x$ on both sides of (\ref{lr-alt}) over any given initial distribution, one obtains an analogous formula for distributed initial conditions.

Note that $\varphi$ and $B_{t}$ are not independent, hence the product of $\varphi$ and the integral over the Brownian motion does not average to zero in general.

\paragraph*{Remark.} 
By formally expanding $M^{-1}_T$ up to second order we get an analogous second order response formula: 
\begin{equation}\label{2nd}
\begin{aligned}
\bE_{\rho_{x}^{\eps}}[f] & \approx \bE_{\rho_{x}^{0}}[\varphi]  + \eps \bE_{\rho_{x}^{0}}\left[\varphi\int_{0}^{T}\xi_{s}\cdot dB_{s} \right]\\  
& + \frac{\eps^{2}}{2} \bE_{\rho_{x}^{0}}\left[\varphi\left\{\left(\int_{0}^{T}\xi_{s}\cdot dW_{s}\right)^2 -\int_0^T |\xi_s|^2\,ds\right\} \right].
\end{aligned}
\end{equation}

@@

We now link our previous considerations with the previous case and consider a nonequilibrium Langevin equation. Specifically, we add a non-gradient perturbation to the drift (\ref{eqn:langevin-1}) of the previous Langevin equation, i.e., we set 
\begin{equation}\label{eqn:langevin-1b}
b(x,t) = (J-R)\nabla H(x) +  B(x) u(t)\,,\quad a(x) = \sqrt{2\beta^{-1}}R^{1/2}
\end{equation}
with the definitions of $J,R$ and $H$ unchanged, $u\in\R^{n}$ being some control variable, and $B(\cdot)\in\R^{2n\times n}$ given by 
\[
B(x) = \left( \begin{array}{c}
0 \\ D(x) \end{array}\right)
\]
The matrix $D(\cdot)\in\R^{n\times n}$ must be chosen such that $B$ satisfies the Fredholm alternative $\mathrm{range}(B(\cdot))\perp \mathrm{ker}((a(\cdot))^{T})$, where $\mathrm{ker}(a^T)$ denotes the kernel of $a^T$.

We are interested in the dynamics under a small perturbation $\delta u$ of the control $u$, Specifically, we assume that $v$ in (\ref{sde2}) is of the form 
\begin{equation}\label{eqn:langevin-1c}
v_{t} = B(X_{t})\delta u(t)
\end{equation}
so that equation (\ref{xi}) that determines the change of measure in terms of the auxiliary control variable $\xi$ (and thus the linear response) reads 
\[
\sqrt{2\beta^{-1}}R^{1/2} \xi_{t} = D(X_{t})\,\delta u_{t}\,.
\]
Hence the equation for $\xi_{t}$ is solvable if and only if $\mathrm{range}(D(\cdot))\perp \mathrm{ker}(R^{1/2})$, which, since $R$ has full rank, means that $D(\cdot)$ must be invertible almost everywhere.

To be more explicit, the equation we are considering has the form
\begin{equation}\label{eqn:langevin-1d}
dX_t =  (J-R)\nabla H(X_{t})dt + B(X_t)(u(t)+\eps \delta u(t))dt+a\, dB_t\,.
\end{equation}
In this case the linear response formula (\ref{lr-alt})  becomes
\begin{equation}\label{lr-alt2}
\lim_{\eps\to 0}\frac{\bE_{\rho_{x}^{\eps}}[\varphi] - \bE_{\rho_{x}^{0}}[\varphi]}{\eps} =  \bE_{\rho_{x}^{0}}\left[\varphi\int_{0}^{T}(\sigma^{-1}D(x_{s})\,\delta u_{s})\cdot dB_{s} \right]\,
\end{equation}
with the shorthands $\sigma=\sqrt{2\beta^{-1}}R^{1/2}$ and 
\[
\varphi = \int_{0}^{T}f(x_{s})ds+g(x_{T})\,.
\]
In other words: 
\begin{equation}\label{eqn:neq-response}
\bE_{\rho_{x}^{\eps}}[\varphi] \approx \bE_{\rho_{x}^0}[\varphi] + \eps\bE_{\rho_{x}^0}\left[\varphi\int_{0}^{T}(\sigma^{-1}D(x_{s})\,\delta u_{s})\cdot dB_{s} \right]\,.
\end{equation}
Here, as before, the expectation on the right hand side of the linear response formula is over all realizations of (\ref{eqn:langevin-1d}) for $\eps=0$ starting at $x$; see also the remark below on the choice of initial conditions.

@@

As one can easily get lost in the various transformations and changes of measure, involving $\nu$, $\mu$, $\rho_{x}^{\eps}$, $\rho^{0}_{x}$ etc. it may be helpful to understand how the linear response formulas (\ref{lr-alt}) or (\ref{lr-alt2}) can be realized algorithmically. To this end, let $(x_0,\,x_{1},\,x_{2},\,x_{3},\ldots)$ with $x_{k}$ be the numerical realization of (\ref{sde2-1}). Let us further suppose that the initial value $x_{0}=x$ is fixed. The simplest possible numerical discretization would be the Euler scheme 
\[
x_{n+1} = x_{n} + \Delta t\, b(x_{n},t_{n}) + \sqrt{\Delta t}\,a(x_{n})\eta_{n+1}\,,\quad x_{0}=x\,, 
\]
with time step $\Delta t=t_{k+1}-t_{k}$ and $\eta_{k}$ i.i.d.~Gaussian random variables with mean 0 and unit covariance. (For a Langevin equation such as (\ref{eqn:langevin-1d}) the Euler scheme is not recommended, but the basic idea stays the same.) Now a simple unbiased estimator of the linear response---i.e. the right hand side in (\ref{lr-alt})---would be 
\begin{equation}\label{lr-est}
\hat{R} = \frac{1}{M}\sum_{i=1}^{M}\left\{f(\{x_{k}(\omega_{i})\}_{0\le k\le N})\sum_{j=0}^{N-1}\hat{\xi}_{i}(\omega_{i})\cdot \eta_{j+1}(\omega_{i})\right\}
\end{equation}
with $N=\lfloor T/\Delta t\rfloor$ and $x_{k}(\omega_{i})$ denoting the $i$-th realization of $x_{k}$, that is generated by the $i$-th realization $(\eta_{1}(\omega_{i}),\ldots,\eta_{k}(\omega_{i}))$ of the Gaussian noise sequence $(\eta_{k})_{k\in\N}$. The time-discrete control variable is given by  
\[
a(x_{k})\hat{\xi}_{k} = v_{t_{k}}
\]
for any given perturbation $v\colon[0,T]\to\R^{m}$.

>>>>

\begin{abstract}
  In this paper, we propose a straightforward generalization of linear response theory on a finite time-horizon to systems in nonequilibrium that are subject to external forcing.  We briefly revisit the standard linear response result for equilibrium systems, where we consider Langevin dynamics as a special case, and then give an alternative derivation using a change-of-measure argument that does not rely on any stationarity or reversibility assumption. This procedure easily enables us to calculate the second order correction to the linear response formula (which may or may not be useful in practice). Furthermore, we outline how the novel nonequilibrium linear response formula can be used to compute optimal controls of molecular systems for cases in which one wants to steer the system to maximize a certain target expectation value. We illustrate our approach with simple numerical examples. 
\end{abstract}

@@

Often, however, one is interested in knowing about the response of the molecular system to perturbation out of equilibrium. The standard linear response formula allows to answer this question, at least partially. In the standard setting it gives us the first order of the change to an equilibrium expectation value as resulting from the nonequilibrium perturbation. Here, first order means first order in the size of the perturbation. This linear response formula has a long history of extensions and generalizations; see, e.g., \cite{marconi2008,ruelle2009} and the references therein. In some sense it has become one of the cornerstones of modern statistical physics since it can be related to the fluctuation dissipation theorem (FDT) which roughly states that for appropriate systems in statistical equilibrium, the average response to small external perturbations can be calculated through the knowledge of suitable correlation functions of the unperturbed statistical system. Linear response formulae on the infinite time-horizon have been shown to hold under rather mild assumptions even for nonequilibrium systems that possess a unique invariant measure \cite{hairer2008}.

There is an increasing number of articles in the literature that report on applications of molecular dynamics to nonequilibrium settings. There are many generalization to so-called nonequilibrium steady states based on the generality of the FDT \cite{Seifert}, but despite its wide use, we are not aware of a linear response formula for equilibrium or nonequilibrium (i.e.~irreversible) systems that does not rely on any kind of stationarity assumption. We will provide such a formula for the case that the underlying dynamics can be described by Langevin dynamics that may not have an invariant measure. Furthermore, we will provide a formula for the second order response of a nonequilibrium Langevin system subject to a small perturbation.

Instead of applying this theory to a molecular system we will go one step further. We will outline how the novel nonequilibrium linear response formula can be used to compute the optimal control of molecular systems. In optimal control one seeks the optimal way to perturb a molecular system, such that a certain target expectation value (e.g. population of certain states) is maximized under constraints on the energy of the control. In general, the control drives the molecular system under control out of equilibrium. Thus, the nonequilibrium linear response formula can be used to find the optimal correction of the present control regarding the expectation value of interest.

@@

In this section, we first want to give a simple and formal derivation of the linear response for equilibrium and nonequilibrium systems. To this end we consider a general It\^o stochastic differential equation of the form
\begin{equation}\label{sde}
dX_{t} = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t}\,,\quad t\ge 0\,,
\end{equation}
where $X_{t}\in\R^{d}$, $b(\cdot,\cdot)$ is a smooth time-dependent vector field, $a(\cdot)$ a smooth field of $d\times n$ matrices and $B_{t}$ is standard Brownian motion in $\R^{n}$. Here $v_{t}\in\R^{d}$ is any given driving force applied to the system, typically an affine function of the form $v_{t}=c(X_{t})u(t)$, with $c(\cdot)\in\R^{d\times k}$ and a bounded measurable function $u\colon[0,\infty)\to\R^{k}$ such that (\ref{sde}) has a strong solution for all $t> 0$ whenever $\eps>0$ is sufficiently small. When (\ref{sde}) without the perturbation $\eps v$ is considered, we write
\begin{equation}\label{sdewo}
dx_{t} = b(x_{t},t)dt +a(x_{t})dB_{t}\,,\quad t\ge 0\,. 
\end{equation}

\subsection{Small perturbations from equilibrium: Langevin dynamics}\label{ssec:equiLRT}

@@

Specifically, we are interested in the case that (\ref{sde}) has the form of a second-order Langevin equation, in which case $d=2n$ and 
\begin{equation}\label{eqn:langevin-1}
b(x,t) = (J-R)\nabla H(x)\,,\quad a = \sqrt{2\beta^{-1}}R^{1/2}\,,
\end{equation}
with $J=-J^{T}$ the canonical $2n\times 2n$ symplectic matrix,
\[
J = \left ( \begin{array}{cc} 
0 & I_{n\times n} \\ -I_{n\times n} &0
\end{array}\right),
\]
$R=R^{T}\ge 0$ the positive semidefinite $2n\times 2n$ friction matrix 
\[
R = \left ( \begin{array}{cc}
0 & 0 \\ 0 & \gamma
\end{array}\right),
\]
for $\gamma\in\R^{n\times n}$ symmetric positive definite, and 
\[
H\colon\R^{2n}\to\R\,,\quad H(x) = \frac{1}{2}|p|^{2} + U(q) \quad (x=(q,p))
\] 
the Hamiltonian of the system. We assume that the potential $U$ is bounded from below. Then, under some mild growth conditions on $U$ for $|q|\to\infty$ (see, e.g., \cite{mattingly2002}), the unperturbed dynamics has a unique invariant measure with density 
\[
\rho_{\infty}(x) = \frac{1}{Z}\exp(-\beta H(x))\,,\quad  Z=\int_{\R^{2n}}  \exp(-\beta H(x))\,dx\,.
\] 
In this case, the linear response result (\ref{lr}) can be recast in form of the better known Green-Kubo formula \cite{risken1996}. Assuming that the forcing is of the form $v_{t}=Cu(t)$ with $u\colon[0,\infty)\to\R^{n}$ and $C\in\R^{2n\times n}$ being any suitable given time-dependent function, the above expression can be recast as 
\begin{equation}\label{GreenKubo}
\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f] - \bE_{\rho^{0}}[f]}{\eps} = \beta\int_{0}^{t}F(t-s)u(s)\,ds
\end{equation}
where 
\[
F(t) = - \int_{\R^{2n}}\bE_{x}[f(x_{t})] C^{T}\nabla H(x)\rho_{\infty}(x)\,dx
\]
is the response function and the expectation under the integral is taken over all realization of the equilibrium process (\ref{sdewo}) starting from $x_{0}=x$. In other words: 
\[
\bE_{\rho^{\eps}}[f] \approx \bE_{\rho_{\infty}}[f] + \eps\beta\int_{0}^{t}F(t-s)u(s)\,ds\,,
\]
where we have used the fact that $\rho^{0}(x,t)= \rho_{\infty}(x)$ for all $t\ge 0$.

@@

\subsubsection*{Girsanov transformation}
We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal2003stochastic}. To this end, let $x_{t}=x_{t}(\omega)$ and $X_{t}=X_{t}(\omega)$ be the solutions of our generic stochastic differential equations
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t}\,,\quad 0\le t\le T  \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t}\,,\quad 0\le t\le T \label{sde2-2}
\end{align}
\end{subequations}
for $T<\infty$ and deterministic initial conditions
\[
x_{0}(\omega) = X_{0}(\omega) = x\quad\textrm{(almost surely)}\,.
\]
Suppose that there exists an auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,.
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to rewrite (\ref{sde2-2}) as
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003stochastic}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{stroock2006}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$. Specifically, let $P$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $Q$ on the space of continuous trajectories by 
\[
dQ=M_{T}dP
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any functional 
\[
\varphi(\omega)=\int_{0}^{T}f((X_{s}(\omega))\,ds + g(X_{T}(\omega))
\] 
that is integrable \wrt $\nu$, we have the identity\footnote{One of the omitted technical details is Novikov's condition \cite[pp.~162]{oksendal2003stochastic} that guarantees that $(M_{t})_{0\le t\le T}$ is a Martingale, which by $\bE_{Q}[1]=\bE_{P}[M_{T}]=1$ implies normalizability of the new probability measure $Q$.}
\[
\bE_{P}[\varphi] := \int_{\Omega} \varphi(\omega)\, dP(\omega) = \int_{\Omega} \varphi(\omega) M_{T}^{-1}(\omega)\,dQ(\omega) =: \bE_{Q}[M^{-1}_{T}\varphi]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)
\]
is the density of $P$ relative to $Q$. Note that the expectation on the right hand side corresponds to the unperturbed dynamics, because $W_{t}$ is a standard Brownian motion under $Q$, and the expectation is over all realizations of (\ref{sde2-3}) starting from $X_{0}=x$. On the other hand, $X_{t}$ under $P$ corresponds to the perturbed dynamics, which should become clear upon comparing equations (\ref{sde2-2}) and (\ref{sde2-3}).

@@

Linearization of $M_{T}^{-1}$ about $\eps=0$, assuming that the control has bounded variance, yields the alternative linear response formula 
\begin{equation}\label{lr-alt}
\lim_{\eps\to 0}\frac{\bE_{P_{x}^{\eps}}[\varphi] + \bE_{P_{x}^{0}}[\varphi]}{\eps} = \bE_{P_{x}^{0}}\left[\varphi \int_{0}^{T}\xi_{s}\cdot dB_{s} \right]\,,
\end{equation}
where, in analogy with the previous linear response formula, $P_{x}^{\eps}$ and $P_{x}^{0}$ denote the path probability measures of perturbed and unperturbed paths---i.e.~the trajectories obtained from (\ref{sde2-2}) and (\ref{sde2-1})---with deterministic initial condition. The path measure $P^{\eps}_{x}$ should not be confused with the phase space density $\rho^{\eps}(x,t)$ in Section \ref{ssec:equiLRT} that solves the Fokker Planck equation. Note moreover that $\varphi$ on the right hand side of (\ref{lr-alt}) is understood as a functional of $(x_{t})_{0\le t\le T}$, i.e.~the solution to (\ref{sde2-1}) with initial condition $x_{0}=x$, with $\xi_{t}$ being the solution to $a(x_{t})\xi_{t}=v_{t}$.

By averaging the the initial values $x$ on both sides of (\ref{lr-alt}) over any given initial distribution, one obtains an analogous formula for distributed initial conditions. Note that $\varphi$ and $B_{t}$ are not independent, hence the product of $\varphi$ and the integral over the Brownian motion does not average to zero in general.

\paragraph*{Remark.} 
By formally expanding $M^{-1}_T$ up to second order we get an analogous second order response formula: 
\begin{equation}\label{2nd}
\begin{aligned}
\bE_{P_{x}^{\eps}}[f] & \approx \bE_{P_{x}^{0}}[\varphi]  + \eps \bE_{P_{x}^{0}}\left[\varphi\int_{0}^{T}\xi_{s}\cdot dB_{s} \right]\\  
& + \frac{\eps^{2}}{2} \bE_{P_{x}^{0}}\left[\varphi\left\{\left(\int_{0}^{T}\xi_{s}\cdot dW_{s}\right)^2 -\int_0^T |\xi_s|^2\,ds\right\} \right].
\end{aligned}
\end{equation}

@@

We now link our previous considerations with the previous case and consider a nonequilibrium Langevin equation. Specifically, we add a non-gradient perturbation to the drift (\ref{eqn:langevin-1}) of the previous Langevin equation, i.e., we set 
\begin{equation}\label{eqn:langevin-1b}
b(x,t) = (J-R)\nabla H(x) +  c(x) u(t)\,,\quad a(x) = \sqrt{2\beta^{-1}}R^{1/2}
\end{equation}
with the definitions of $J,R$ and $H$ unchanged, $u\in\R^{n}$ being some bounded measurable control, and $c(\cdot)\in\R^{2n\times n}$ given by 
\[
c(x) = \left( \begin{array}{c}
0 \\ D(x) \end{array}\right)
\]
The matrix $D(\cdot)\in\R^{n\times n}$ must be chosen such that $c$ satisfies the Fredholm alternative $\mathrm{range}(c(\cdot))\perp \mathrm{ker}((a(\cdot))^{T})$, where $\mathrm{ker}(a^T)$ denotes the kernel of $a^T$.

We are interested in the dynamics under a small perturbation $\delta u$ of the control $u$, Specifically, we assume that $v$ in (\ref{sde2}) is of the form 
\begin{equation}\label{eqn:langevin-1c}
v_{t} = c(X_{t})\delta u(t)
\end{equation}
so that equation (\ref{xi}) that determines the change of measure in terms of the auxiliary control variable $\xi$ (and thus the linear response) reads 
\[
\sqrt{2\beta^{-1}}R^{1/2} \xi_{t} = D(X_{t})\,\delta u(t)\,.
\]
Hence the equation for $\xi_{t}$ is solvable if and only if $\mathrm{range}(D(\cdot))\perp \mathrm{ker}(R^{1/2})$, which, since $R$ has full rank, means that $D(\cdot)$ must be invertible almost everywhere.

To be more specific, the equation we are considering has the form
\begin{equation}\label{eqn:langevin-1d}
dX_t =  (J-R)\nabla H(X_{t})dt + c(X_t)(u(t)+\eps \delta u(t))dt+a\, dB_t\,.
\end{equation}
In this case the linear response formula (\ref{lr-alt})  becomes
\begin{equation}\label{lr-alt2}
\lim_{\eps\to 0}\frac{\bE_{P_{x}^{\eps}}[\varphi] - \bE_{P_{x}^{0}}[\varphi]}{\eps} =  \bE_{P_{x}^{0}}\left[\varphi\int_{0}^{T}(\sigma^{-1}D(x_{s})\,\delta u(s))\cdot dB_{s} \right]\,
\end{equation}
with the shorthands $\sigma=\sqrt{2\beta^{-1}}R^{1/2}$ and 
\[
\varphi = \int_{0}^{T}f(x_{s})ds+g(x_{T})\,.
\]
In other words: 
\begin{equation}\label{eqn:neq-response}
\bE_{P_{x}^{\eps}}[\varphi] \approx \bE_{P_{x}^0}[\varphi] + \eps\bE_{P_{x}^0}\left[\varphi\int_{0}^{T}(\sigma^{-1}D(x_{s})\,\delta u(s))\cdot dB_{s} \right]\,.
\end{equation}
Here, as before, the expectation on the right hand side of the linear response formula is over all realizations of (\ref{eqn:langevin-1d}) for $\eps=0$ starting at $x$; see also the remark below on the choice of initial conditions.

@@

As one can easily get lost in the various transformations, involving path measures $P$, $Q$, $P_{x}^{\eps}$, $P^{0}_{x}$ or phase space densities $\rho^{\eps}$, $\rho^{0}$, $\rho_{\infty}$, it may be helpful to understand how the linear response formulas (\ref{lr-alt}) or (\ref{lr-alt2}) can be realized algorithmically. To this end, let $(x_0,\,x_{1},\,x_{2},\,x_{3},\ldots)$ with $x_{k}$ be the numerical realization of (\ref{sde2-1}). Let us further suppose that the initial value $x_{0}=x$ is fixed. The simplest possible numerical discretization would be the Euler scheme 
\[
x_{n+1} = x_{n} + \Delta t\, b(x_{n},t_{n}) + \sqrt{\Delta t}\,a(x_{n})\eta_{n+1}\,,\quad x_{0}=x\,, 
\]
with time step $\Delta t=t_{k+1}-t_{k}$ and $\eta_{k}$ i.i.d.~Gaussian random variables with mean 0 and unit covariance. (For a Langevin equation such as (\ref{eqn:langevin-1d}) the Euler scheme is not recommended, but the basic idea stays the same.) Now a simple unbiased estimator of the linear response---i.e. the right hand side in (\ref{lr-alt})---would be 
\begin{equation}\label{lr-est}
\hat{R} = \frac{1}{M}\sum_{i=1}^{M}\left\{\varphi(\{x_{k}(\omega_{i})\}_{0\le k\le N})\sum_{j=0}^{N-1}\hat{\xi}_{i}(\omega_{i})\cdot \eta_{j+1}(\omega_{i})\right\}
\end{equation}
with $N=\lfloor T/\Delta t\rfloor$ and $x_{k}(\omega_{i})$ denoting the $i$-th realization of $x_{k}$, that is generated by the $i$-th realization $(\eta_{1}(\omega_{i}),\ldots,\eta_{k}(\omega_{i}))$ of the Gaussian noise sequence $(\eta_{k})_{k\in\N}$. The time-discrete control variable is given by  
\[
a(x_{k})\hat{\xi}_{k} = v_{t_{k}}
\]
for any given perturbation $v\colon[0,T]\to\R^{m}$.
